---
title: "09 — GATK Hard Filter QC"
author: "Natalie Love"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

# Overview

This notebook evaluates the GATK hard filters applied in script `08_filter_snps.sh` to the
joint-genotyped *Gymnocladus dioicus* SNP callset (61 WGS samples, 30×, Hap 1 reference).

**Goals:**

1. Confirm the QUAL30 filter is correctly expressed (0 sites removed — is this to be expected or is something up?)
2. Plot annotation value distributions (PASS vs. failed) to evaluate whether filter thresholds are placed at natural valleys in the data
3. Check the Ti/Tv ratio as an independent quality metric
4. Identify which filters are doing the most work and flag any that may need adjustment

**Input files (produced by `08_filter_snps.sh`):**

- Tagged VCF (all sites, filter labels applied):
  `~/kct_genomics/output_files/gatk_filter_hap1/gymno_hap1.snps.tagged.vcf.gz`
- Filtered VCF (PASS only):
  `~/kct_genomics/output_files/gatk_filter_hap1/gymno_hap1.snps.filtered.vcf.gz`

---

# Step 1 — Extract annotation values (run on Curie)

The following bash chunks are run on the Curie server, not locally. They extract per-site
annotation values from the tagged VCF into a flat TSV for plotting in R, and run
`bcftools stats` on the filtered VCF for Ti/Tv evaluation.



**1a. Extract annotations from tagged VCF**
bcftools is a command-line toolkit for reading, writing, and manipulating VCF/BCF files
 ┌──────────┬───────────────────────────────────────────────────┐                                                                                                       
  │  Token   │                   What it pulls                   │
  ├──────────┼───────────────────────────────────────────────────┤
  │ %CHROM   │ Chromosome                                        │
  ├──────────┼───────────────────────────────────────────────────┤
  │ %POS     │ Position                                          │                                                                                                       
  ├──────────┼───────────────────────────────────────────────────┤
  │ %QUAL    │ Phred-scaled variant quality score                │                                                                                                       
  ├──────────┼───────────────────────────────────────────────────┤                                                                                                       
  │ %FILTER  │ The filter label (e.g. PASS, QD2, QD2;SOR3)       │
  ├──────────┼───────────────────────────────────────────────────┤                                                                                                       
  │ %INFO/QD │ A field from the INFO column named QD             │
  ├──────────┼───────────────────────────────────────────────────┤                                                                                                       
  │ %INFO/FS │ Same pattern — INFO field FS                      │
  ├──────────┼───────────────────────────────────────────────────┤                                                                                                       
  │ ...      │ (same for SOR, MQ, MQRankSum, ReadPosRankSum, DP) │
  ├──────────┼───────────────────────────────────────────────────┤                                                                                                       
  │ \t       │ Tab separator between fields                      │
  ├──────────┼───────────────────────────────────────────────────┤                                                                                                       
  │ \n       │ Newline at the end of each site's row             │
  └──────────┴───────────────────────────────────────────────────┘   

```{bash eval=FALSE}
# Run on Curie — conda activate gatk
# Extracts one row per SNP site with PASS/FAIL label, all 7 annotation values, and INFO/DP.
# MQRankSum and ReadPosRankSum will be '.' (missing) at homozygous sites — expected.
# INFO/DP is total depth summed across all 61 samples; used to preview script 10 DP thresholds.
# Output: ~16M rows, ~600 MB uncompressed.

bcftools query \
  -H \
  -f '%CHROM\t%POS\t%QUAL\t%FILTER\t%INFO/QD\t%INFO/FS\t%INFO/SOR\t%INFO/MQ\t%INFO/MQRankSum\t%INFO/ReadPosRankSum\t%INFO/DP\n' \
  ~/kct_genomics/output_files/gatk_filter_hap1/gymno_hap1.snps.tagged.vcf.gz \
  > ~/kct_genomics/output_files/gatk_filter_hap1/snp_annotations.tsv
```

**1b. Run bcftools stats on filtered VCF (for Ti/Tv)**
Done: 04/18/2026
```{bash eval=FALSE}
# Run on Curie — conda activate gatk
# Computes summary statistics including Ti/Tv ratio on the PASS-only VCF.

bcftools stats \
  ~/kct_genomics/output_files/gatk_filter_hap1/gymno_hap1.snps.filtered.vcf.gz \
  > ~/kct_genomics/output_files/gatk_filter_hap1/bcftools_stats_filtered.txt
```

**1c. Download output files to local machine**

```{bash eval=FALSE}
# Run locally — copy annotation table and stats from Curie
scp nlove@10.2.0.53:~/kct_genomics/output_files/gatk_filter_hap1/snp_annotations.tsv \
  ../output_data/snp_annotations.tsv

scp nlove@10.2.0.53:~/kct_genomics/output_files/gatk_filter_hap1/bcftools_stats_filtered.txt \
  ../output_data/bcftools_stats_filtered.txt
```

---

# Step 1 Setup

```{r setup, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)

# set wd to where the Rmd is
knitr::opts_knit$set(root.dir = here::here())

# Path to annotation table downloaded from Curie (stored in output_data/, gitignored)
annotations_path <- "output_data/snp_annotations.tsv"
stats_path       <- "output_data/bcftools_stats_filtered.txt"

# Run identifier — change this when re-running after pipeline changes
# e.g. "original" for the first run, "mq30_run" after adding 04b MAPQ filter
run_id <- "original"

# Output paths for comparison CSVs
out_filter_counts <- "output_data/filter_counts.csv"
out_qc_summary    <- "output_data/run_qc_summary.csv"
```

---

# Step 2 — Load annotation table
```{r load-data, message=FALSE}
ann <- read_tsv(
  annotations_path,
  col_names = c("CHROM", "POS", "QUAL", "FILTER",
                "QD", "FS", "SOR", "MQ", "MQRankSum", "ReadPosRankSum", "DP"),
  comment   = "#",      # bcftools -H adds a header line starting with #
  na        = ".",      # missing values (MQRankSum/ReadPosRankSum at hom sites)
  show_col_types = FALSE
) |>
  slice_sample(n = 1e6) |>   # 1M rows sufficient for distribution shape; saves RAM
  mutate(pass = if_else(FILTER == "PASS", "PASS", "FAIL"))

cat("Sampled sites:", nrow(ann), "\n")
cat("PASS:", sum(ann$pass == "PASS"), "\n")
cat("FAIL:", sum(ann$pass == "FAIL"), "\n")
```

---

# Step 3 — QUAL distribution

**QUAL Meaning**
 QUAL is GATK's Phred-scaled confidence that a variant exists at that site at all — across all 61 samples combined. Specifically:                     
                                                            
  - QUAL = 30 means 99.9% confident a real variant is there                                                                                            
  - QUAL = 60 means 99.9999% confident                      
  - It's calculated from the likelihood of the observed reads given a variant vs. no variant

**QUAL filtering**
QUAL30 removed 0 sites in script 08. Two explanations:

1.  **All sites genuinely have QUAL ≥ 30** — with 61 samples at 30× WGS, joint genotyping
   produces high QUAL scores across the board, so QUAL < 30 is rare or absent - we do indeed have no sites less than 30.

Plotting the QUAL distribution distinguishes these: if the expression is misspelled we'd
see values below 30 in the PASS sites; if it's working correctly, the distribution should
be entirely ≥ 30.

NL Notes: Yes, all QUAL scores are above 30, no PASS below 30

```{r qual-dist, fig.width=8, fig.height=4}

ann %>%
  ggplot(aes(x = QUAL, fill = pass)) +
  geom_histogram(bins = 100, alpha = 0.7, position = "identity") +
  # plot vertical line at QUAL=30
  geom_vline(xintercept = 30, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(values = c("PASS" = "#2196F3", "FAIL" = "#F44336")) +
  # modify x axis to zoom in and out
  scale_x_continuous(limits = c(0, 100)) +
  labs(title = "QUAL distribution — all SNP sites",
       x = "QUAL (Phred-scaled variant quality)",
       y = "Number of sites",
       fill = NULL) +
  theme_bw()
```

**Interpretation:** If all PASS sites are well above the red dashed line (QUAL = 30), the
filter expression is correct and the 0-removal result is genuine. If PASS sites extend below
30, the expression may be silently failing.

NL: Checked!
---

# Step 4 — Read depth (DP) distribution

INFO/DP is the total read depth summed across all 61 samples at each site. This is a preview
for setting site-level DP thresholds in script 10 (population-level filtering).

- **Low-DP tail** (e.g. < ~200 for 61 samples × ~30×): likely poorly-covered or misaligned
  regions; genotypes unreliable. A common rule of thumb is to require mean DP ≥ ~5–10× per
  sample, so min INFO/DP ≈ n_samples × min_per_sample.
  Tail at 0.99?
- **High-DP tail**: likely repetitive or copy-number-variable regions where reads from
  multiple loci pile up. Typically flagged above mean + 2–3 SD, or a round number like 2× mean.
  
NL Notes:
Average summed depth:

```{r dp-distribution, fig.width=10, fig.height=4, message=FALSE, warning=FALSE}
min_depth <- 5
targed_depth <- 30
n_samples <- 61

dp_summary <- ann %>%
  filter(!is.na(DP)) %>%
  summarise(
    mean_dp   = mean(DP),
    median_dp = median(DP),
    sd_dp     = sd(DP),
    p01       = quantile(DP, 0.01),
    p99       = quantile(DP, 0.99)
  )

cat(sprintf(
  "DP summary (PASS+FAIL sites):\n  Mean: %.0f  Median: %.0f  SD: %.0f\n  1st pct: %.0f  99th pct: %.0f\n",
  dp_summary$mean_dp, dp_summary$median_dp, dp_summary$sd_dp,
  dp_summary$p01, dp_summary$p99
))

# Suggested thresholds: min = average 5 depth, max = median depth * 2
# median 
dp_min_suggested <- min_depth*n_samples
dp_max_suggested <- dp_summary$median_dp*2

# filter SNPs that are NA or have depth not between 1 and 99 percentile
# for visualizing
dp_filter <- ann |>
  filter(!is.na(DP), DP >= dp_summary$p01, DP <= dp_summary$p99)

ggplot(data = dp_filter, aes(x = DP, fill = pass)) +
  geom_histogram(bins = 100, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = dp_min_suggested, linetype = "dashed",
             color = "orange", linewidth = 0.8) +
  geom_vline(xintercept = dp_max_suggested, linetype = "dashed",
             color = "orange", linewidth = 0.8) +
  annotate("text", x = dp_min_suggested, y = Inf,
           label = sprintf("average\ndepth 5\n(%.0f)", dp_min_suggested),
           vjust = 2, hjust = 1.1, color = "orange", size = 3) +
  annotate("text", x = dp_max_suggested, y = Inf,
           label = sprintf("2x median depth\n(%.0f)", dp_max_suggested),
           vjust = 2, hjust = -0.1, color = "orange", size = 3) +
  scale_fill_manual(values = c("PASS" = "#2196F3", "FAIL" = "#F44336")) +
  labs(title = "INFO/DP distribution — total depth across all 61 samples",
       subtitle = "Tails clipped at 1 and 99 percentile | Orange lines = suggested script 10 thresholds (average depth of 5-60 per SNP)",
       x = "INFO/DP (total depth, all samples)", y = "Number of sites", fill = NULL) +
  theme_bw()
```

**Interpretation:** NL: No natural values to guide cutoffs, but we will now filter out high tail reads (likely repetative regions) - could consider tightening upper to 3000 if we need to

---

# Step 5 — Annotation distributions (PASS vs. FAIL, excluding DP)

For each annotation, we plot overlapping histograms of PASS (blue) and FAIL (red) sites.
A well-placed threshold should sit at a clear valley between the two distributions.

**QD — Quality by Depth** *(fails if < 2.0; 1,363,777 sites removed)*

QD normalizes the variant quality score (QUAL) by the total read depth at that site. Without this normalization, high-coverage sites accumulate high QUAL scores simply because more reads vote for the variant — even if each individual read provides weak evidence. QD corrects for this: a site with QUAL=100 and depth=10 has QD=10 (strong), while QUAL=100 and depth=200 has QD=0.5 (weak). Low QD indicates a variant call that looks confident overall but is actually poorly supported per read — the hallmark of a sequencing error at a well-covered site.

**FS — Fisher Strand Bias** *(fails if > 60.0; 75,218 sites removed)*

FS uses Fisher's exact test to detect strand bias — the tendency for a variant to be supported predominantly by reads on one DNA strand (forward or reverse) rather than both. Real variants should appear equally on both strands; artifacts often don't because they arise from strand-specific sequencing errors or library preparation issues. FS of 0 = perfectly balanced; higher values = more biased. The threshold of 60 is intentionally lenient for WGS data — SOR (below) is the primary strand bias filter.

**SOR — Strand Odds Ratio** *(fails if > 3.0; 518,874 sites removed)*

SOR is an improved strand bias metric that addresses a weakness in FS: FS tends to penalize high-coverage sites even when bias is mild, because large sample sizes make small imbalances statistically significant. SOR uses an odds ratio formulation that is less sensitive to coverage depth, making it more reliable for WGS data. Values near 0 = no bias; higher values = increasing bias.

**MQ — Mapping Quality** *(fails if < 40.0; 801,726 sites removed)*

MQ is the root mean square of the mapping quality scores of all reads covering the site. Mapping quality reflects how confidently each read was placed in the genome by the aligner (BWA). Low MQ means reads at that site could plausibly have come from elsewhere in the genome — common in repetitive regions, paralogs, or poorly assembled scaffolds where reads map ambiguously. MQ=60 is the maximum for BWA-MEM (uniquely placed read); MQ=0 means the read maps equally well to multiple locations.

**MQRankSum — Mapping Quality Rank Sum Test** *(fails if < −12.5; 305 sites removed)*

MQRankSum compares the mapping quality of reads supporting the reference allele vs. reads supporting the alternate allele using a Wilcoxon rank-sum test. At a real variant, ref and alt reads should have similar mapping quality. A strongly negative value means alt-supporting reads map significantly worse than ref-supporting reads — suggesting the alt allele is an artifact caused by misaligned reads rather than a true variant. Only calculated at heterozygous sites (requires both ref and alt reads).

**ReadPosRankSum — Read Position Rank Sum Test** *(fails if < −8.0; 4 sites removed)*

ReadPosRankSum tests whether alt allele reads tend to occur at the ends of sequencing reads rather than in the middle. Bases near read ends are sequenced with lower accuracy, so variants called predominantly from end-of-read positions are likely artifacts. A strongly negative value = alt allele reads are systematically near read ends. Like MQRankSum, only calculated at het sites, and very few sites fail this filter in high-quality WGS data.

```{r annotation-distributions, message=FALSE, warning=FALSE}
thresholds <- tibble(
  annotation = c("QD", "FS", "SOR", "MQ", "MQRankSum", "ReadPosRankSum"),
  threshold  = c(2.0,  60.0, 3.0,  40.0, -12.5,         -8.0)
)

ann_long <- ann |>
  select(pass, QD, FS, SOR, MQ, MQRankSum, ReadPosRankSum) |>
  pivot_longer(-pass, names_to = "annotation", values_to = "value") |>
  filter(!is.na(value))

ann_long |>
  group_by(annotation) |>
  mutate(lo = quantile(value, 0.005, na.rm = TRUE),
         hi = quantile(value, 0.995, na.rm = TRUE)) |>
  filter(value >= lo, value <= hi) |>
  ungroup() |>
  ggplot(aes(x = value, fill = pass)) +
  geom_histogram(bins = 100, alpha = 0.7, position = "identity") +
  geom_vline(data = thresholds,
             aes(xintercept = threshold),
             linetype = "dashed", color = "red", linewidth = 0.7) +
  scale_fill_manual(values = c("PASS" = "#2196F3", "FAIL" = "#F44336")) +
  facet_wrap(~annotation, scales = "free", ncol = 2) +
  labs(title = "Annotation distributions — PASS vs. FAIL sites",
       subtitle = "Red dashed line = current hard filter threshold | Tails clipped at 0.5th/99.5th percentile",
       x = "Annotation value", y = "Number of sites", fill = NULL) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))

rm(ann_long); gc()
```

**What to look for:**

- **QD:** FAIL sites (red) should be concentrated at values < 2. If the red peak extends
  far right of the threshold, some real variants may be caught.
- **FS / SOR:** FAIL sites should cluster at high values (right tail). Low FS/SOR values
  are expected to be PASS.
- **MQ:** FAIL sites should cluster at low values (< 40). A large FAIL peak may indicate
  many repetitive/poorly-mapped regions in the reference.
- **MQRankSum / ReadPosRankSum:** These have many missing values (only calculated at het
  sites). Very few sites should fail these filters given the low counts in the log.

NL
                                                                                                                       
  - QD < 2.0 — the FAIL peak was cleanly concentrated below the threshold with good separation from PASS. Well-placed.  
  - FS > 60.0 — lenient but intentional. SOR is doing the real strand bias work. No reason to tighten.
  - SOR > 3.0 — removed 518K sites, the second biggest filter. Distribution looked appropriate.                         
  - MQ < 40.0 — the spike at MQ=60 confirms most sites map cleanly; the sites below 40 are genuinely problematic        
  repetitive regions.                                                                                                   
  - MQRankSum < −12.5 and ReadPosRankSum < −8.0 — only 305 and 4 sites removed respectively. These are so conservative  
  they're basically just catching extreme outliers, which is fine.                                                      
                                                            
  The Ti/Tv of 2.17 is the strongest argument for leaving things alone — that's right in the middle of the expected     
  2.0–2.2 range for clean WGS data, meaning the filters are removing artifacts without over-filtering real variants.

---

# Step 6 — Per-filter failure counts

From the script 08 log (recorded here for reference):

```{r filter-counts}
raw_snps  <- 16021590
pass_snps <- 13672054

filter_counts <- tibble(
  filter      = c("QD2", "QUAL30", "FS60", "SOR3", "MQ40",
                  "MQRankSum-12.5", "ReadPosRankSum-8"),
  sites_failed = c(1363777, 0, 75218, 518874, 801726, 305, 4),
  threshold   = c("QD < 2.0", "QUAL < 30.0", "FS > 60.0", "SOR > 3.0",
                  "MQ < 40.0", "MQRankSum < -12.5", "ReadPosRankSum < -8.0")
) |>
  mutate(pct_of_total = round(sites_failed / raw_snps * 100, 2))

filter_counts |>
  arrange(desc(sites_failed)) |>
  knitr::kable(
    col.names = c("Filter", "Sites failed", "Expression", "% of total SNPs"),
    format.args = list(big.mark = ",")
  )

# Export — append to CSV so multiple runs accumulate
filter_counts |>
  mutate(run = run_id, raw_snps = raw_snps) |>
  select(run, everything()) |>
  write_csv(out_filter_counts,
            append = file.exists(out_filter_counts))

cat("Filter counts written to:", out_filter_counts, "\n")
```

```{r filter-counts-plot, fig.width=7, fig.height=4}
filter_counts |>
  mutate(filter = fct_reorder(filter, sites_failed)) |>
  ggplot(aes(x = filter, y = sites_failed)) +
  geom_col(fill = "#F44336", alpha = 0.8) +
  geom_text(aes(label = scales::comma(sites_failed)),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.2))) +
  labs(title = "Sites failed per filter",
       subtitle = "Total SNPs before filtering: 16,021,590 | After: 13,672,054",
       x = NULL, y = "Sites failed") +
  theme_bw()
```

---

# Step 7 — Ti/Tv ratio

Transitions (Ti) are purine↔purine or pyrimidine↔pyrimidine substitutions (A↔G, C↔T).
Transversions (Tv) are purine↔pyrimidine substitutions (A↔C, A↔T, G↔C, G↔T).

In the genome, transitions occur more often than transversions. For WGS data, the expected
Ti/Tv ratio is approximately **2.0–2.2** for the whole genome. Values well below 2.0 suggest
an excess of false positives (artifacts tend to be transversions); values well above 2.2 may
indicate over-filtering.

```{r titv, message=FALSE}
# Parse Ti/Tv from bcftools stats output
# The TSTV line format is: TSTV  <id>  <ts>  <tv>  <ts/tv>  <ts (1st ALT)>  <tv (1st ALT)>  <ts/tv (1st ALT)>
# [2] is a record ID (0 = all samples); actual values start at [3]
stats_lines <- read_lines(stats_path)

titv_line <- stats_lines[str_starts(stats_lines, "TSTV")]

titv_ti <- NA_real_; titv_tv <- NA_real_; titv_ratio <- NA_real_

if (length(titv_line) > 0) {
  titv_vals  <- str_split(titv_line, "\t")[[1]]
  titv_ti    <- as.numeric(titv_vals[3])
  titv_tv    <- as.numeric(titv_vals[4])
  titv_ratio <- as.numeric(titv_vals[5])
  tibble(
    Metric = c("Transitions (Ti)", "Transversions (Tv)", "Ti/Tv ratio"),
    Value  = c(
      formatC(titv_ti, format = "f", digits = 0, big.mark = ","),
      formatC(titv_tv, format = "f", digits = 0, big.mark = ","),
      as.character(titv_ratio)
    )
  ) |> knitr::kable()
} else {
  cat("TSTV line not found in bcftools stats output. Check stats file format.\n")
}
```

**Interpretation:** Ti/Tv of ~2.0–2.2 indicates a well-filtered callset. If Ti/Tv is below
2.0, consider relaxing the most aggressive filter (currently QD2 at 1.36M sites).

NL: the Ti/Tv is 2.17, which is good
---

# Step 8 — Summary and threshold recommendations

```{r summary}
cat("=== Script 08 filter summary ===\n")
cat("Biallelic SNPs before filtering: 16,021,590\n")
cat("PASS SNPs after hard filters:    13,672,054\n")
cat("Sites removed:                    2,349,536 (14.7%)\n\n")
cat("Dominant filter: QD2 (1,363,777 sites = 8.5% of total)\n")
cat("Second: MQ40 (801,726 sites = 5.0%)\n")
cat("Third:  SOR3 (518,874 sites = 3.2%)\n\n")
cat("Note: sites can fail multiple filters, so counts above are not additive.\n")
```

Based on the plots above, record any threshold adjustments recommended here before
proceeding to population-level filtering (script 10).

| Annotation | Current threshold | Recommended adjustment | Rationale |
|------------|------------------|----------------------|-----------|
| QD | < 2.0 | TBD | |
| QUAL | < 30.0 | TBD | |
| FS | > 60.0 | TBD | |
| SOR | > 3.0 | TBD | |
| MQ | < 40.0 | TBD | |
| MQRankSum | < -12.5 | TBD | |
| ReadPosRankSum | < -8.0 | TBD | |

---

# Step 8b — Export run QC summary

```{r export-qc-summary}
# Assemble one-row summary for this run and append to comparison CSV
qc_summary <- tibble(
  run           = run_id,
  date          = as.character(Sys.Date()),
  n_samples     = 61,
  raw_snps      = raw_snps,
  pass_snps     = pass_snps,
  pct_removed   = round((raw_snps - pass_snps) / raw_snps * 100, 2),
  titv          = titv_ratio,
  titv_ti       = titv_ti,
  titv_tv       = titv_tv,
  dp_mean       = round(dp_summary$mean_dp, 1),
  dp_median     = round(dp_summary$median_dp, 1),
  dp_sd         = round(dp_summary$sd_dp, 1),
  dp_min_filter = dp_min_suggested,
  dp_max_filter = dp_max_suggested
)

qc_summary |> knitr::kable()

write_csv(qc_summary,
          out_qc_summary,
          append = file.exists(out_qc_summary))

cat("QC summary written to:", out_qc_summary, "\n")
```

---

# Step 9 — SNP density across chromosomes

Visualizes where SNPs fall along each main chromosome, both before and after hard filtering.
Gaps indicate centromeres, low-complexity regions, or assembly gaps. Unplaced scaffolds
(not matching `chr##` pattern) are excluded.

```{r snp-density, fig.width=12, fig.height=16, message=FALSE, warning=FALSE}
ann |>
  filter(str_detect(CHROM, "^chr\\d+$")) |>
  mutate(window = floor(POS / 1e6)) |>
  count(CHROM, window, pass) |>
  ggplot(aes(x = window, y = n, fill = pass)) +
  geom_col(width = 1, alpha = 0.8) +
  scale_fill_manual(values = c("PASS" = "#2196F3", "FAIL" = "#F44336")) +
  facet_wrap(~CHROM, scales = "free_x", ncol = 2) +
  labs(title = "SNP density across chromosomes — PASS vs. filtered sites",
       x = "Position (Mb)", y = "SNPs per 1 Mb window", fill = NULL) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

---

# Session info

```{r session-info}
sessionInfo()
```
